Assessing \Htwo{} and \COtwo{} Network Strategies in Europe

### 2023-12-08 report/submissions/journal/paper/paper.tex
paper appendix: add cost comparison for half-price co2 net

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While the \modBase{} has neither a \COtwo{} nor a \Htwo{} network, the \modHybrid{} is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\COtwo{} Infrastructure'' combines transport, storage and sequestration, and ``\Htwo{} Infrastructure'' combines transport and storage. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

% In summary, the \COtwo{} network enables the transport of carbon from the lowest-cost CC sources to CU and CS carbon sinks, reducing the need for high-cost DAC.
% The creation of a \Htwo{} network enables the cost-effective delivery of hydrogen from regions with a surplus of low-cost renewable energy to high-demand industrial centers with exogenous hydrogen demand and a surplus of low-cost CC potentials, reducing the need for DAC and for high-cost electrolysis at the industrial centers.
% This highlights how both the implementation of a \COtwo{} or a \Htwo{} network unlocks low-cost carbon capture potentials at geographically fixed point sources across the continent.

@@

As we have also discussed above, the \modCO{} captures large amounts of low-cost carbon from bioenergy and industrial activities that were untapped in the \modBase{}, especially in Central and Eastern Europe, and the \COtwo{} network smoothes out price differences for captured carbon across regions (see Fig.~\ref{fig:balance_map}, top left). Hydrogen prices still vary strongly, with low price ``valleys'' in the regions with good renewable resources such as the Iberian Peninsula or Denmark, and high price ``peaks'' in Central Europe (see bottom left). The model optimally locates CU in the regions where it can minimize costs for its material inputs, carbon and hydrogen. Thus, when carbon prices are smoothed out in the \modCO{}, CU moves to the hydrogen price valleys in the periphery of the continent. The captured carbon is transported from Central Europe through several large pipeline corridors to these locations to produce carbonaceous fuels.
Similarly to CU, the model optimally locates CS in regions where it can minimize costs for its only material input, carbon. Thus, CS moves to the carbon price valleys (which are not strongly pronounced due to the \COtwo{} network, but still prevalent) among the subset of regions with access to offshore sequestration sites. In the \modCO, these are sites  sites in the North Sea (Northern Germany, Norway, Southeastern England), Baltic Sea (Latvia, Eastern Denmark) and Mediterranean Sea (Italy, Greece). CU and CS compete for carbon sources and thus there is some interaction for their optimal placement.
Finally, in the absence of an \Htwo{} network, the system places some electrolysis and SMR in Central Europe to supply exogenous hydrogen demand for industrial processes and fuel cell land transport (see bottom left). Due to the additional electricity demand for electrolysis, the model switches some winter heating from heat pumps to gas\=/fueled boilers and CHPs, and thus emphasizes overall more gas imports into Central Europe.

The \Htwo{} network on the other hand, smoothes out hydrogen prices across the entire continent (see bottom right). Thus, the model chooses to locate CU in the carbon price valleys across Western, Central and Eastern Europe (see top right). Low-cost hydrogen from regions with abundant renewable resources is transported to these regions with low-price carbon from abundant sources of fossil, biogenic and process emissions for point-source CC. The \Htwo{} network is also used to supply the exogenous \Htwo{} demand in these regions for industry and fuel cell land transport at much lower costs than local electrolysis or SMR.
The model locates CS again in the regions that represent carbon price valleys among all those with access to offshore sequestration sites. In the case of the \modH{}, due to the lack of a \COtwo{} network all CC potentials from point sources are exhausted in all offshore sequestration regions, and the carbon is then sequestered. The additionally needed carbon to reach 200~Mt of CS is captured via DAC in the regions offshore sequestration regions with the lowest costs for DAC, with the largest share deployed in Portugal (see top right).

The results illustrate the commonalities and differences between the two grids: both a \COtwo{} network and an \Htwo{} network prefer distributed CC at point sources and electrolysis in regions with abundant renewable resources, choosing the lowest-cost production sites for each while trading off transport costs. Nonetheless, the \COtwo{} network is forced to deploy some electrolysis in regions with poor renewables to supply their exogenous hydrogen demands for industry and fuel cell land transport.

The differences between the CO2-Grid and H2-Grid in the context of CU, CS and DAC are notable and highlight the unique operational characteristics of each grid type.
CU always chooses the sites with the lowest costs for its hydrogen and carbon inputs. With the \COtwo{} network, \COtwo{} costs are levelized across the continent, and CU is thus located in the regions with the lowest \Htwo{} costs. With the \Htwo{} network, \Htwo{} costs are levelized across the continent, and CU is thus located in the regions with the lowest \COtwo{} costs. Since regions with low-cost \COtwo{} in the \modH{} are very spread out across the continent, the CU locations are also more distributed, while in the \modCO{} they cluster in a few pockets of low-cost \Htwo{}.
CS co-locates in both models in the regions with the lowest \COtwo{} costs among those with offshore sequestration potential. However, due to the lack of \COtwo{} transport in the \modH{}, additional CS is located in the offshore locations where DAC, and therefore electricity, can be produced at the lowest costs.
Finally, DAC in the \modCO{} is only used for CU, and is thus co-located in the regions with a mix of the lowest electricity costs (for running DAC to produce \COtwo{}, the first CU input material), and the lowest hydrogen costs for producing \Htwo{} (the second CU input material). These regions happen to co-incide with where the best renewable resources are located. In the \modH{}, DAC is used for both CU and CS and is located in the regions with the lowest costs for electricity to run the DAC, with the additional requirement of DAC for CS having to be placed in regions with offshore sequestration access.

In summary, both network models provide largely similar distributions of electrolysis and CC sites, but very different siting of CU, CS and DAC due to their distinct regional price differentiations for \COtwo{} and \Htwo{}. Both systems unlock captured carbon potentials in Central Europe. The system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} needs to deploy electrolysis powered by poor renewable resources as well as additional gas infrastructure in Central Europe.

@@

The \modHybrid{} combines the advantages of hydrogen and carbon networks. While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role. Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of the two grids in the \modHybrid{}. Note that despite similar overall capacities, the investment in the hydrogen infrastructure is four times the investment in the carbon infrastructure.

@@

\begin{figure}
    \centering
    \includegraphics*[width=0.8\linewidth]{comparison/emission-reduction/figures/90_nodes/objective_heatmap.png}
    \caption{Total annual system cost for the different network models and \COtwo{} targets. While the \modBase{} has neither a \COtwo{} nor a \Htwo{} network, the \modHybrid{} is allowed to expand both.}
    \label{fig:objective_heatmap}
\end{figure}

@@

% [Dynamics \modCO{} vs \modH{} (old text blocks)]
% There are four notable dynamics or differences between the two grid strategies:
% First, the production of FT fuels always moves to where low-cost carbon and low-cost hydrogen can be brought together. In the \modCO{ }, carbon is moved from Western Europe to the Iberian Peninsula, the British Isles and to Denmark to produce FT fuels centrally in few regions. In the \modH{}, hydrogen is transported in the opposite direction to produce FT fuels decentrally across all regions in Central Europe with locally captured \COtwo{} from point sources.
% Second, the \modCO{ } captures more low-cost carbon from biomass, especially in Central and Eastern Europe, and less high-cost carbon from DAC in Greece and Southern Italy. Instead, the \COtwo{} network transports carbon from Central and Eastern Europe to Greece and Italy, where it is sequestered, or feeds into CU together with locally produced, low-cost electrolytic hydrogen. The \modH{} directly operates electrolyzers and DAC in Greece and Southern Italy, with electricity from low-cost renewables, and thus avoids significant network buildouts in this region.
% Third, sequestration in the \modH{} occurs more decentralized, spread out across several regions with access to offshore sequestration sites, making use of point-source CC potentials in each of these regions. In the \modCO{ }, sequestration is clustered at fewer sequestration sinks, each collecting low-cost \COtwo{} from neighboring regions. However, there is no centralization at only one or two sites, since all the sequestration sites offer the same costs and the system minimizes system-wide carbon transport distances to about eight to ten sequestration sites.
% And fourth, how the three types of CU are spread across the continent is driven by the costs of carbon and hydrogen (the latter mostly driven by the costs of renewable electricity) as well as by renewables capacity factors: Clearly, all carbonaceous fuel production requires sites where both carbon and hydrogen are available, and they moreover prefer sites with low carbon and hydrogen prices. Based on our model assumptions, FT fuel synthesis has very limited flexibility and needs to be run alost at baseload (ramp-down only possible to 90\% of maximum load). Thus, for FT fuel synthesis sites in Northern Europe with more electricity from wind power are preferred to sites in Southern Europe, powered mostly by solar. Methanathion and Methanolisation both have higher flexibility, being able to ramp down to 50\% load during the night, for example. Thus, these fuel production sites prefer the best renewable resource sites in Southern Europe. And finally, as methanation produces synthetic methane that needs to be fed into the gas grid, methanation prefers those sites with good interconnection to the existing gas network. Since Italy possesses a large gas network with easy interconnection to Central Europe, methanation prefers sites in Southern Italy (and Greece, which would be connected to Southern Italy) over the Iberian Peninsula, where Methanolisation is more prevalent.

@@

% Therefore, one can conclude that a \COtwo{} network favors a decentralized CC and a CU centralized at regions with abundant renewable resources. A \Htwo{} network favors a centralized electrolysis system at regions with abundant renewable resources and a decentralized CU system.

% Together with understanding the dynamics described above, we can highlight three cost advantages of the \Htwo{} network in contrast to the \COtwo{} network: Capturing carbon at the lowest-cost regions reduces 4 billion euros in spending on DAC. However, the \modH{} produces electrolytic hydrogen in the lowest-cost regions and in total spends 7 billion euros less on renewable energy (14 billion euros less on solar, and 7 billion euros more on wind) as well as 1 billion euros less on SMR. Furthermore, without a \Htwo{} network, additional \Htwo{} storage is needed in Spain at a cost of 5 billion euros to store hydrogen produced during the summer for utilization in FT fuel production during the winter.
% And finally, the \modCO{ } spends 7 billion euros more on gas boilers, gas plants, gas infrastructure, and methanation.
% TODO: why??
% The costs of the respective networks, on the other hand, are almost equal: the \modCO{ } spends 8 billion euros on the \COtwo{} network, while the \modH{} even spends 10 billion euros on the \Htwo{} network. These factors add up to account for 14 billion euros of the difference of 20 billion euros in system costs between the two scenarios (785 billion euros for the \COgrid{} and 765 billion euros for the \modH{}, see Fig.~\ref{fig:objective_heatmap}).

>>>>

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While the \modBase{} has neither a \COtwo{} nor an \Htwo{} network, the \modHybrid{} is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\COtwo{} Infrastructure'' combines transport, storage and sequestration, and ``\Htwo{} Infrastructure'' combines transport and storage. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

% In summary, the \COtwo{} network enables the transport of carbon from the lowest-cost CC sources to CU and CS carbon sinks, reducing the need for high-cost DAC.
% The creation of an \Htwo{} network enables the cost-effective delivery of hydrogen from regions with a surplus of low-cost renewable energy to high-demand industrial centers with exogenous hydrogen demand and a surplus of low-cost CC potentials, reducing the need for DAC and for high-cost electrolysis at the industrial centers.
% This highlights how both the implementation of a \COtwo{} or an \Htwo{} network unlocks low-cost carbon capture potentials at geographically fixed point sources across the continent.

@@

As we have also discussed above, the \modCO{} captures large amounts of low-cost carbon from bioenergy and industrial activities that were untapped in the \modBase{}, especially in Central and Eastern Europe, and almost none of which is used locally (see Fig.~\ref{fig:balance_map}, top left). A large part of the captured carbon is transported from Western Europe to the Iberian Peninsula, the British Isles and Denmark, where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part of captured carbon is transported to methanation plants in Italy and Greece, where the synthetic methane output is fed into the well-established gas grid since Italy serves as a large gas import hub for Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. In absence of an \Htwo{} network, the system places most electrolysis in regions with the best renewable resources to supply local CU, and smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe. The model also emphasizes gas imports into Central Europe. Hydrogen prices still vary strongly across the continent, with low price ``valleys'' in regions with good renewable resources such as the Iberian Peninsula or Denmark, and high price ``peaks'' in Central Europe (see bottom left).

To understand the dynamics of the \modCO{}, we observe that the system locates CU in the regions where it can minimize costs for its material inputs, carbon and hydrogen. As the \COtwo{} network largely smoothes out carbon prices across the continent, CU is located in the hydrogen price valleys on the periphery of the continent. Similarly, the model locates CS in regions where it can minimize costs for its only material input, carbon. Thus, CS is located in the carbon price valleys among the subset of regions with access to offshore sequestration sites. Due to the \COtwo{} network these valleys are not strongly pronounced, but still prevalent and impact CS locations. DAC is only used for CU and thus co-located in regions with a mix of low electricity costs (for running DAC) and low hydrogen costs (where CU is located, see above).

The \Htwo{} network on the other hand, transports low-cost hydrogen from regions with abundant renewable resources to regions with low-price carbon from abundant sources of fossil, biogenic and process emissions (see bottom right). The main transportation routes go from Spain and the United Kingdom to Central Europe to produce synfuels with locally captured carbon from point sources. Also, the \Htwo{} network supplies exogenous \Htwo{} demand for industry in Central Europe at much lower costs than local electrolysis or SMR.

To explain the dynamics of the \modH{}, we observe that the model locates CU in the carbon price valleys across Western, Central and Eastern Europe (see top right), as the \Htwo{} network smoothes out hydrogen prices across the entire continent (see bottom right). CS is again located in the carbon price valleys among those regions with access to offshore sequestration sites. Due to the lack of a \COtwo{} network the full CC potentials from point sources are exhausted in all offshore sequestration regions, and the carbon is then sequestered. The additionally needed carbon to reach 200~Mt of CS is captured via DAC in the offshore sequestration regions with the lowest costs for DAC, with the largest share deployed in Portugal (see top right). DAC is used for both CU and CS in this model, and for both purposes DAC is located in the regions with the lowest costs for electricity.

In summary, both network models prefer distributed CC at point sources and electrolysis in regions with abundant renewable resources, choosing the lowest-cost production sites for each while trading off transport costs. However, differences between the two models are notable for the placement of CU, CS and DAC due to their distinct \COtwo{} and \Htwo{} price patterns. CU locations are more distributed in the \modH{}, while they cluster in a few pockets of low-cost \Htwo{} in the \modCO{}. CS co-locates in the regions with the lowest \COtwo{} costs (among those with offshore sequestration potential), but these regions are very different between the models. DAC is placed in regions with the best renewable resources in both models, but it is used for only CU in the \modCO{} and for both CU and CS in the \modH{}. Finally, the system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy some electrolysis in regions with poor renewables resources as well as additional gas infrastructure in Central Europe.

@@

The \modHybrid{} combines the advantages of hydrogen and carbon networks. While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role. Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of the two networks in the \modHybrid{}. Note that despite similar overall capacities, the investment in the hydrogen infrastructure is four times the investment in the carbon infrastructure.

@@

\begin{figure}
    \centering
    \includegraphics*[width=0.8\linewidth]{comparison/emission-reduction/figures/90_nodes/objective_heatmap.png}
    \caption{Total annual system cost for the different network models and \COtwo{} targets. While the \modBase{} has neither a \COtwo{} nor an \Htwo{} network, the \modHybrid{} is allowed to expand both.}
    \label{fig:objective_heatmap}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.9\linewidth]{difference/comparison/subsidy/figures/90_nodes/cost_bar.png}
    \caption[short]{Net change in system cost when reducing assumed costs for \carbon{} pipelines by 50\%.}
    \label{fig:net-negative_cost_bar}
\end{figure}

% [Dynamics \modCO{} vs \modH{} (old text blocks)]
% There are four notable dynamics or differences between the two network strategies:
% First, the production of FT fuels always moves to where low-cost carbon and low-cost hydrogen can be brought together. In the \modCO{ }, carbon is moved from Western Europe to the Iberian Peninsula, the British Isles and to Denmark to produce FT fuels centrally in few regions. In the \modH{}, hydrogen is transported in the opposite direction to produce FT fuels decentrally across all regions in Central Europe with locally captured \COtwo{} from point sources.
% Second, the \modCO{ } captures more low-cost carbon from biomass, especially in Central and Eastern Europe, and less high-cost carbon from DAC in Greece and Southern Italy. Instead, the \COtwo{} network transports carbon from Central and Eastern Europe to Greece and Italy, where it is sequestered, or feeds into CU together with locally produced, low-cost electrolytic hydrogen. The \modH{} directly operates electrolyzers and DAC in Greece and Southern Italy, with electricity from low-cost renewables, and thus avoids significant network buildouts in this region.
% Third, sequestration in the \modH{} occurs more decentralized, spread out across several regions with access to offshore sequestration sites, making use of point-source CC potentials in each of these regions. In the \modCO{ }, sequestration is clustered at fewer sequestration sinks, each collecting low-cost \COtwo{} from neighboring regions. However, there is no centralization at only one or two sites, since all the sequestration sites offer the same costs and the system minimizes system-wide carbon transport distances to about eight to ten sequestration sites.
% And fourth, how the three types of CU are spread across the continent is driven by the costs of carbon and hydrogen (the latter mostly driven by the costs of renewable electricity) as well as by renewables capacity factors: Clearly, all carbonaceous fuel production requires sites where both carbon and hydrogen are available, and they moreover prefer sites with low carbon and hydrogen prices. Based on our model assumptions, FT fuel synthesis has very limited flexibility and needs to be run alost at baseload (ramp-down only possible to 90\% of maximum load). Thus, for FT fuel synthesis sites in Northern Europe with more electricity from wind power are preferred to sites in Southern Europe, powered mostly by solar. Methanathion and Methanolisation both have higher flexibility, being able to ramp down to 50\% load during the night, for example. Thus, these fuel production sites prefer the best renewable resource sites in Southern Europe. And finally, as methanation produces synthetic methane that needs to be fed into the gas grid, methanation prefers those sites with good interconnection to the existing gas network. Since Italy possesses a large gas network with easy interconnection to Central Europe, methanation prefers sites in Southern Italy (and Greece, which would be connected to Southern Italy) over the Iberian Peninsula, where Methanolisation is more prevalent.

@@

% Therefore, one can conclude that a \COtwo{} network favors a decentralized CC and a CU centralized at regions with abundant renewable resources. An \Htwo{} network favors a centralized electrolysis system at regions with abundant renewable resources and a decentralized CU system.

% Together with understanding the dynamics described above, we can highlight three cost advantages of the \Htwo{} network in contrast to the \COtwo{} network: Capturing carbon at the lowest-cost regions reduces 4 billion euros in spending on DAC. However, the \modH{} produces electrolytic hydrogen in the lowest-cost regions and in total spends 7 billion euros less on renewable energy (14 billion euros less on solar, and 7 billion euros more on wind) as well as 1 billion euros less on SMR. Furthermore, without an \Htwo{} network, additional \Htwo{} storage is needed in Spain at a cost of 5 billion euros to store hydrogen produced during the summer for utilization in FT fuel production during the winter.
% And finally, the \modCO{ } spends 7 billion euros more on gas boilers, gas plants, gas infrastructure, and methanation.
% TODO: why??
% The costs of the respective networks, on the other hand, are almost equal: the \modCO{ } spends 8 billion euros on the \COtwo{} network, while the \modH{} even spends 10 billion euros on the \Htwo{} network. These factors add up to account for 14 billion euros of the difference of 20 billion euros in system costs between the two scenarios (785 billion euros for the \COgrid{} and 765 billion euros for the \modH{}, see Fig.~\ref{fig:objective_heatmap}).
